---
title: "svdvis: Visualizing SVD, PCA, and related methods"
author: |
  | Neo Christopher Chung
  | nchchung@gmail.com
date: "`r Sys.Date()`"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{svdvis Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This package provides several visualization functions for singular value decomposition (SVD), principal component analysis (PCA), factor analysis (FA), logistic factor analysis (LFA), and other related methods.

## Simulated data
To use in this vignette, we create a simulated dataset, with `m=500` variables (rows) and `n=20` samples (columns). Particularly, it contains a latent variable that resembles a case-control study. After applying SVD to the datasets, we also name the rows and the columns of the right singular vectors `svd.obj$v` for labels in visualization.

```{r, fig.show='hold'}
set.seed(1234)
library(svdvis)
B = c(runif(100, min=0, max=1), rep(0,400))
L = c(rep(1, 10), rep(-1, 10))
L = L / sd(L)
E = matrix(rnorm(500*20), nrow=500)
Y = B %*% t(L) + E

svd.obj = svd(Y)
colnames(svd.obj$v) = paste0("V",1:20)
rownames(svd.obj$v) = paste0("Sample",1:20)
```

In this setup, a few right singular vectors contained in `svd.obj$v` may capture systematic variation in the observed data `Y`. Since the right singular vectors are ordered according to the singular values in a descending order, the top (or first) `r` right singular vectors refers to `svd.obj$v[,1:r]`. Note that principal components (PCs) can be obtained by multiplying singular values `svd.obj$d` and right singular vectors `svd.obj$v`. All examples in this vignette and all functions in `svdvis` can utilize `weights="sv"` to quickly visualize PCs.

## Scree plot

A scree plot visualizes percentages of variance explained by singular vectors in a descending order. `svd.scree` is simply a wrapper function using `ggplot2`. In high-dimensional datasets, the number of points in a scree plot may be too large. It may be good to look at a subset of singular values. You can specify `subr` in `svd.scree` function, which "zooms in" to the top `subr` singular values.

```{r}
svd.scree(svd.obj, subr=5,
          axis.title.x="Full scree plot", axis.title.y="% Var Explained")
```

Note that if `subr` is not specified, one full-sized scree plot is returned.

## Paired scatterplots

Scatter plots are often utilized to look at the top 2 right singular vectors. `svd.scatter` produces a matrix of scatterplots of all pairs among `r` right singular vectors.  
```{r}
svd.scatter(svd.obj)
```

The above plot crams in too many pairs. We can specify `r` to visualize only the top `r` right singular vectors. In this example, additional arguments such as `group` and `alpha` are included:
```{r}
svd.scatter(svd.obj, r=3, alpha=.5,
            group=c(rep("Group 1", 10), rep("Group 2", 10)))
```

## Heat map
Let's create a heat map of the top `r=5` right singular vectors:
```{r}
svd.heatmap(svd.obj, r=5)
```


## Parallel coordinates plot
A parallel coordinates plot shows `r` dimensions in `r` parallel lines, which are equally spaced. All data points are rescaled to (0,1) and the top `r` singular vectors are visualized from left to right. Different groups are colored accordingly: 

```{r}
svd.parallel(svd.obj, r=5, alpha=.5,
             group=c(rep("Group 1", 10), rep("Group 2", 10)))
```

## Radial coordinates plot
A radial coordinates plot visualize `r` dimensions in a circle, around where `r` anchors are placed. Each of `n` vectors is mapped onto a circle, using its data as spring constants. Prior to mapping, each column is rescaled to have numeric values between 0 and 1.

```{r}
svd.radial(svd.obj, r=3,
           group=c(rep("Group 1", 10), rep("Group 2", 10)))
```

## Tips and remarks
All functions in `svdvis` use `ggplot2`. Therefore, the visual output can be saved and modified in a conventional manner. Feel free to experiment the source codes for more complex or interesting cases.

While this vignette focused on using the results of SVD, an optional argument `weights="sv"` can be used for visualizing PCs. Note that `weights="sv"` is simply calling `weights = svd.obj$d[1:r]`. 
  
Outputs from other dimension reduction methods can be used. Provide the `r` vectors to `svd.obj` in any function. Note that the input must be a `n * r` matrix that contains `r` vectors as columns. An optional argument `group` can be used to differentially indicate `n` samples (points, lines, etc).

For example, logistic factor analysis captures population structure from a large and diverse set of genome sequences and is related to SVD and PCA. A R package `lfa` computes `r` logistic factors, as columns. You can easily make a parallel coordinates plot (and others) by `svd.parallel(svd.obj=lfa(genotypes, 10))`.
